---
title: "project #2: Merging data to create a US election map"

execute: 
  warning: false
  message: false
format:
  html:
    page-layout: full
    title-block-banner: true
    code-fold: true 
---

### Now let's load some election data and map data and plot it:

```{r}
library(sf)
library(ggplot2)
library(leaflet)
library(tidyverse)
library(ggthemes)
library(ggnewscale)
library(ggpubr)
library(highcharter)

load("data/dataverse_shareable_presidential_county_returns_1868_2020.Rdata")

#Citation dataset: Algara, Carlos; Sharif Amlani, 2021, "Replication Data for: Partisanship & Nationalization in American Elections: Evidence from Presidential, Senatorial, & Gubernatorial Elections in the U.S. Counties, 1872-2020", https://doi.org/10.7910/DVN/DGUMFI, Harvard Dataverse, V1, UNF:6:glfQoiLzpXDGTfErebfBIQ== [fileUNF]

#Citation File: Algara, Carlos; Sharif Amlani, 2021, "Replication Data for: Partisanship & Nationalization in American Elections: Evidence from Presidential, Senatorial, & Gubernatorial Elections in the U.S. Counties, 1872-2020", https://doi.org/10.7910/DVN/DGUMFI, Harvard Dataverse, V1; dataverse_shareable_gubernatorial_county_returns_1865_2020.Rdata [fileName]

us_data2<- st_read("data/gz_2010_us_050_00_5m.json", quiet=TRUE)


#replace non-printable characters with blank space
us_data2$NAME <- str_replace_all(us_data2$NAME, "[^[:print:]]", "")
```

#### Let's try to make the map and election data as similar as possible: exclude to only counties in the continental US. {style="color:teal;"}

```{r}
#these are the state codes for hawaii, alaska, and other territories

#Let's try to make the county data as similar as possible: exclude to continental US.

us_data2<- us_data2 %>% filter(!STATE %in% c("02", "15", "60", "66", "69", "72", "78", "74"))
pres_elections_release<- pres_elections_release %>% filter(!sfips %in% c("02", "15", "60", "66", "69", "72", "78", "74"))

```

#### Let's first begin the merge process by first checking if every county in map data only has one unique identifier/code. Next, we should have 39 codes for each county in the election data, because there have been 39 election cycles included in the data (1868-2020). {style="color:teal;"}

##### Since there historically has been less counties, we will filter the data to start in the 1916 election. This is because 1912 is the first election year that the entire continental US was included. However, in 1912 there were three candidates and the "Progressive" party came in second. Since 1916, Democrats and Republicans have been the two largest parties in the elections. Because our data only includes Democratic and Republican votes, I've chosen to start from 1916 for these reasons. {style="color:cornflowerblue;"}

```{r}
#checking map data 
us_data2<- st_make_valid(us_data2)
count<- us_data2 %>% count(GEO_ID)
#every single county (3109) has one entry- yay!! 

#double check: 

unique_count<- us_data2 %>% pull(GEO_ID) %>% 
  unique() %>% length()
#3109 unique geo-ids, which is same length of dataset. Yay! 

#now move onto the next dataset: since we have 39 election cycles, we should have 39 of each entry for each election 1868-2020.
count2<- pres_elections_release %>% count(fips)

#We clearly do not have 39s for every value. How many 39s do i have? 
count2 %>% count(n) %>% view()

#View how many counties were in each election year
county_summary<- pres_elections_release %>% group_by(election_year) %>% summarize(num_counties=n_distinct(fips))

#Historically, 1912 was the first year that there were all 48 states (excluding hawaii and alaska). 
#1912 was a weird election year with former president teddy roosevelt running as a progressive, placing second. Since 1914, Democrat and Republicans have been two biggest parties.

##NOTE: there are some years where the majority of the state voted third party, this will make intensity of vote for Republican/Democrat appear less intense. 


#Let's filter data to include election cycles 1916-2020 
dataafter1916<- pres_elections_release %>% filter(election_year >= 1916)

#recount, now we should have 27 entries for each year. Lots of missing data. 

count2<- dataafter1916 %>% count(fips)

#how many 27s do we have 
count2 %>% count(n) %>% view()

#now move onto the next dataset: should be 39 of each entry for each election 1868-2020
count2<- pres_elections_release %>% count(fips)
#how many 39s do i have 
count2 %>% count(n) %>% view()

county_summary<- pres_elections_release %>% group_by(election_year) %>% summarize(num_counties=n_distinct(fips))
#1912 was the first year that there were all 48 states (excluding hawaii and alaska)
#1912 was weird election year with former president teddy roosevelt running as a progressive, placing second. Start from 1914 where democrat and republican have been biggest two parties since. 

##NOTE: there are some years where the majority of the state voted third party, this will make the difference between republican and democrat appear white. 

dataafter1916<- pres_elections_release %>% filter(election_year >= 1916)

#recount, now should have 27 entries. Lots of missing data. 

count2<- dataafter1916 %>% count(fips)

#how many 27s do you have 
count2 %>% count(n) %>% view()


```

#### Even starting from 1916, we do not have 27 entries for each election year. This is because many counties in 2020 did not historically exist. However, we still want to include this data. The next step is to make a full data set with every county 27 times, even if it does not include data. {style="color:teal;"}

```{r}
#doesn't include missing year data, we want to add missing data so we can fill with NA

#want to create year and fips columns and fill in our missing data with NA

#all distinct fips codes, pull into a vector 
fips_codes<- dataafter1916 %>% distinct(fips) %>% pull()
#sequence every 4 years from 1912-2020
election_years<- seq(1916, 2020, by=4)
#create data frame for all combinations of distinct fips and election years 
complete_data<- expand.grid(fips= fips_codes, election_year= election_years) 

#compares complete data with original data, returns rows that are not in original data
missing_data<- complete_data %>% anti_join(dataafter1916, by=c("fips", "election_year"))


#combines the missing rows with the original dataset

#bind the missing rows with the original dataset
newelectiondata<- bind_rows(dataafter1916, missing_data)

#check counts again, should have 27 of each

count3<- newelectiondata %>% count(fips)
count3 %>% count(n) %>% view()

                                           
```

#### Now every county has 27 entries- we are good! There are 3125 (not including NA) counties which isn't that much more than our other data set (3109). Let's see which ones don't overlap. {style="color:teal;'"}

##### First, we need to make a column to compare the counties in the data sets on. "GEO_ID" in our map data is just "00500000US" + the fips code that exists in our election data set. Using that, we can easily make a matching GEO_ID in our election data. {style="color:cornflowerblue;"}

```{r}
#make new column GEOID to match the us_data2 county identifier 
newelectiondata<- newelectiondata %>%  mutate(GEO_ID= paste0("0500000US", fips))


```

#### Now we will check if every GEO_ID in our election data is in our map data and vice versa, identifying the ones which aren't. {style="color:teal;"}

```{r}
#| results: hide
#is every unique GEO_ID in newelectiondata in us_data2? 
(newelectiondata$GEO_ID %>% unique) %in% (us_data2$GEO_ID %>% unique)

#name every unique ID which is not in us_data2, 17 values
missingnewelectiondata<- (newelectiondata$GEO_ID %>% unique)[!(newelectiondata$GEO_ID %>% unique) %in% (us_data2$GEO_ID %>% unique)]


#make a dataset to show election data not in map data 
missing_geo_ids<- newelectiondata %>% filter(GEO_ID %in% missingnewelectiondata)

#do some research into the counties:


#make a dataset to show years
missing_geo_ids<- newelectiondata %>% filter(GEO_ID %in% missingnewelectiondata)


#this is dade, florida (changed to miami dade after 1996 election); campbell, georgia (ceased to exist after 1928 election); Milton, GA (merged with Fulton after 1932); Ormsby, NV (county until 1969); armstrong, SD (county until 1952); ogala lakota, sd (renamed county after 2014); washabaugh, sd (merged after 1976 election); washington sd (until 1943); elizabeth city, va (merged in 1952); nansemond, va (extinct after 1972); norfolk, va (changed to chesapeake in 1963); princess anne, va (merged in 1963); warwick, va (merged 1952); clifton forge, va (merged w allegheny in 2001); south boston city, va (merged in 1996); south norfolk city, va (beccame chesapeake with norfolk in 1963)


#check other way around 
(us_data2$GEO_ID %>% unique) %in% (newelectiondata$GEO_ID %>% unique)

#check which values aren't in election data
missingusdata2<- (us_data2$GEO_ID %>% unique)[!(us_data2$GEO_ID %>% unique) %in% (newelectiondata$GEO_ID %>% unique)]

#none!

```

#### Our check has shown that every map GEO_ID is in election data, but 17 election data GEO_ID's are not in the map data. After some research, I have found that some counties have simply changed their names and either the new name or old name is in the map data set. (dade-\>miami dade, shannon-\>ogala lakota, ormsby-\>carson city). We will change these names/ID's in the election data to reflect which name/ID is in the map data. {style="color:teal;"}

##### Second, some counties (Campbell GA, Washington SD) have ceased to exist or merged into several counties, making it hard to include in the data. We will simply drop these counties. {style="color:cornflowerblue;"}

```{r}
#single name changes
newelectiondata<- newelectiondata %>% mutate(GEO_ID=ifelse(GEO_ID=="0500000US12025", "0500000US12086", GEO_ID)) %>% mutate(GEO_ID=ifelse(county_name=="OGALA LAKOTA", "0500000US46113", GEO_ID)) %>% mutate(GEO_ID= ifelse(county_name=="ORMSBY", "0500000US32510", GEO_ID))

#drop no-longer existing or merged into several different counties
newelectiondata<- newelectiondata %>% filter(!GEO_ID %in% c("0500000US13041", "0500000US46133"))
```

#### For the rest of the counties, they have merged together in later years. For these, we will take the years that both of the counties exist and add the vote totals together, changing the name and ID to the merged county. We will start with Milton and Fulton, which is now just Fulton, GA. {style="color:teal;"}

```{r}
#take the years that milton exists, change GEOID/rename to Fulton, and add the Milton & Fulton values together 
milton <- newelectiondata %>% filter(GEO_ID %in% c("0500000US13203", "0500000US13121") & election_year %in% c(1916, 1920, 1924, 1928)) %>% mutate(GEO_ID= ifelse(county_name=="MILTON",  "0500000US13121", GEO_ID), county_name= ifelse(county_name=="MILTON", "FULTON", county_name)) %>%
  group_by(county_name, election_year, GEO_ID) %>% 
  summarize(across(c(democratic_raw_votes, republican_raw_votes, pres_raw_county_vote_totals_two_party), sum, na.rm=TRUE), .groups= 'drop')

#filter out Milton and Fulton in relevant years 
miltonfultonout <- newelectiondata %>%
  filter(!(GEO_ID %in% c("0500000US13203", "0500000US13121") & election_year %in% c(1916, 1920, 1924, 1928)))

#bind new data into the set with Milton/Fulton filtered out in relevant years 
newelectiondata <- bind_rows(milton, miltonfultonout)

```

#### Repeat the process for Armstrong and Dewey, now just Dewey, SD. {style="color:teal;"}

```{r}
#repeat process for armstrong->dewey county, SD

armstrong <- newelectiondata %>% filter(GEO_ID %in% c("0500000US46001", "0500000US46041") & election_year %in% c(1928, 1932, 1936, 1940, 1944, 1948, 1952)) %>%
  mutate(GEO_ID= ifelse(county_name=="ARMSTRONG" & state=="SD", "0500000US46041", GEO_ID), county_name= ifelse(county_name=="ARMSTRONG" & state=="SD", "DEWEY", county_name)) %>%
  group_by(county_name, election_year, GEO_ID) %>% 
  summarize(across(c(democratic_raw_votes, republican_raw_votes, pres_raw_county_vote_totals_two_party), sum, na.rm=TRUE), .groups= 'drop')

#filter out dewey and armstrong in relevant years 
deweyfilterout <- newelectiondata %>%
  filter(!(GEO_ID %in% c("0500000US46001", "0500000US46041") & election_year %in% c(1928, 1932, 1936, 1940, 1944, 1948, 1952)))

#bind new data into the set with Armstrong/Dewey filtered out in relevant years 
newelectiondata <- bind_rows(armstrong, deweyfilterout)
```

#### Repeat the process for South Norfolk City and Norfolk, which are now together Chesapeake, VA. {style="color:teal;"}

```{r}
#combine south norfolk city, va and norfolk, va to become chesapeake 
norfolk <- newelectiondata %>% filter(GEO_ID %in% c("0500000US51129", "0500000US51785") & election_year %in% c(1916, 1920, 1924, 1928, 1932, 1936, 1940, 1944, 1948, 1952, 1956, 1960)) %>%
  mutate(GEO_ID= ifelse(GEO_ID=="0500000US51129", "0500000US51550", GEO_ID), county_name= ifelse(county_name=="NORFOLK" & state=="VA", "CHESAPEAKE", county_name), GEO_ID= ifelse(GEO_ID=="0500000US51785", "0500000US51550", GEO_ID), county_name= ifelse(county_name=="SOUTH NORFOLK CITY" & state=="VA", "CHESAPEAKE", county_name)) %>%
  group_by(county_name, election_year, GEO_ID) %>% 
  summarize(across(c(democratic_raw_votes, republican_raw_votes, pres_raw_county_vote_totals_two_party), sum, na.rm=TRUE), .groups= 'drop')

#filter out norfolks in relevant years 
norfolkfilterout <- newelectiondata %>%
  filter(!(GEO_ID %in% c("0500000US51129", "0500000US51785") & election_year %in% c(1916, 1920, 1924, 1928, 1932, 1936, 1940, 1944, 1948, 1952, 1956, 1960)))

#bind new data into the set with norfolks filtered out in relevant years 
newelectiondata <- bind_rows(norfolk, norfolkfilterout)

```

#### Repeat the process for South Boston City and Halifax, now just Halifax, VA. {style="color:teal;"}

```{r}
#south boston city-> halifax, va (merged in 1996); 
southbostoncity <- newelectiondata %>% filter(GEO_ID %in% c("0500000US51780", "0500000US51083") & election_year %in% c(1960, 1964, 1968, 1972, 1976, 1980, 1984, 1988, 1992)) %>%
  mutate(GEO_ID= ifelse(county_name=="SOUTH BOSTON CITY" | county_name=="SOUTH BOSTON", "0500000US51083", GEO_ID), county_name= ifelse(county_name=="SOUTH BOSTON CITY" | county_name=="SOUTH BOSTON", "HALIFAX", county_name)) %>%
  group_by(county_name, election_year, GEO_ID) %>% 
  summarize(across(c(democratic_raw_votes, republican_raw_votes, pres_raw_county_vote_totals_two_party), sum, na.rm=TRUE), .groups= 'drop')

#filter out sbc in relevant years 
sbcfilterout <- newelectiondata %>%
  filter(!(GEO_ID %in% c("0500000US51780", "0500000US51083") & election_year %in% c(1960, 1964, 1968, 1972, 1976, 1980, 1984, 1988, 1992)))

#bind new data into the set with SBC/halifax filtered out in relevant years 
newelectiondata <- bind_rows(southbostoncity, sbcfilterout)

```

#### Repeat process for Clifton Forge and Alleghany, now just Alleghany, VA. {style="color:teal;"}

```{r}
#Clifton forge va-> alleghany 
cliftonforge <- newelectiondata %>% filter(GEO_ID %in% c("0500000US51560", "0500000US51005") & election_year %in% c(1916, 1920, 1924, 1928, 1932, 1936, 1940, 1944, 1948, 1952, 1956, 1960, 1964, 1968, 1972, 1976, 1980, 1984, 1988, 1992, 1996, 2000)) %>%
  mutate(GEO_ID= ifelse(county_name=="CLIFTON FORGE" | county_name=="CLIFTON FORGE CITY", "0500000US51005", GEO_ID), county_name= ifelse(county_name=="CLIFTON FORGE" | county_name=="CLIFTON FORGE CITY", "ALLEGHANY", county_name)) %>%
  group_by(county_name, election_year, GEO_ID) %>% 
  summarize(across(c(democratic_raw_votes, republican_raw_votes, pres_raw_county_vote_totals_two_party), sum, na.rm=TRUE), .groups= 'drop')

#filter out in relevant years 
clifforgeout <- newelectiondata %>%
  filter(!(GEO_ID %in% c("0500000US51560", "0500000US51005") & election_year %in% c(1916, 1920, 1924, 1928, 1932, 1936, 1940, 1944, 1948, 1952, 1956, 1960, 1964, 1968, 1972, 1976, 1980, 1984, 1988, 1992, 1996, 2000)))

#bind new data into the set with clifton forge/alleghany filtered out in relevant years 
newelectiondata <- bind_rows(cliftonforge, clifforgeout)

```

#### Repeat the process for Warwick and Newport News, now just Newport News, VA. {style="color:teal;"}

```{r}
#warwick, va -> newport news
warwick <- newelectiondata %>% filter(GEO_ID %in% c("0500000US51189", "0500000US51700") & election_year %in% c(1916, 1920, 1924, 1928, 1932, 1936, 1940, 1944, 1948)) %>%
  mutate(GEO_ID= ifelse(county_name=="WARWICK", "0500000US51700", GEO_ID), county_name= ifelse(county_name=="WARWICK", "NEWPORT NEWS", county_name)) %>%
  group_by(county_name, election_year, GEO_ID) %>% 
  summarize(across(c(democratic_raw_votes, republican_raw_votes, pres_raw_county_vote_totals_two_party), sum, na.rm=TRUE), .groups= 'drop')


#filter out in relevant years 
warwickout <- newelectiondata %>%
  filter(!(GEO_ID %in% c("0500000US51189", "0500000US51700") & election_year %in% c(1916, 1920, 1924, 1928, 1932, 1936, 1940, 1944, 1948)))

#bind new data into the set with warwick/nn filtered out in relevant years 
newelectiondata <- bind_rows(warwick, warwickout)


```

#### Repeat the process for Nansemond and Suffolk, now just Suffolk, VA. {style="color:teal;"}

```{r}
#nansemond-> suffolk va 

nansemond <- newelectiondata %>% filter(GEO_ID %in% c("0500000US51123", "0500000US51800") & election_year %in% c(1916, 1920, 1924, 1928, 1932, 1936, 1940, 1944, 1948, 1952, 1956, 1960, 1964, 1968)) %>%
  mutate(GEO_ID= ifelse(county_name=="NANSEMOND", "0500000US51800", GEO_ID), county_name= ifelse(county_name=="NANSEMOND", "SUFFOLK", county_name)) %>%
  group_by(county_name, election_year, GEO_ID) %>% 
  summarize(across(c(democratic_raw_votes, republican_raw_votes, pres_raw_county_vote_totals_two_party), sum, na.rm=TRUE), .groups= 'drop')


#filter out in relevant years 
nansout <- newelectiondata %>%
  filter(!(GEO_ID %in% c("0500000US51123", "0500000US51800") & election_year %in% c(1916, 1920, 1924, 1928, 1932, 1936, 1940, 1944, 1948, 1952, 1956, 1960, 1964, 1968)))

#bind new data into the set with nan/suffolk filtered out in relevant years 
newelectiondata <- bind_rows(nansemond, nansout)

```

#### Repeat the process for Elizabeth City and Hampton, now just Hampton, VA. {style="color:teal;"}

```{r}
#elizabeth city-> hampton 

elizabethcity <- newelectiondata %>% filter(GEO_ID %in% c("0500000US51055", "0500000US51650") & election_year %in% c(1916, 1920, 1924, 1928, 1932, 1936, 1940, 1944, 1948)) %>%
  mutate(GEO_ID= ifelse(county_name=="ELIZABETH CITY", "0500000US51650", GEO_ID), county_name= ifelse(county_name=="ELIZABETH CITY", "HAMPTON", county_name)) %>%
  group_by(county_name, election_year, GEO_ID) %>% 
  summarize(across(c(democratic_raw_votes, republican_raw_votes, pres_raw_county_vote_totals_two_party), sum, na.rm=TRUE), .groups= 'drop')

#filter out in relevant years 
elizabethcityout <- newelectiondata %>%
  filter(!(GEO_ID %in% c("0500000US51055", "0500000US51650") & election_year %in% c(1916, 1920, 1924, 1928, 1932, 1936, 1940, 1944, 1948)))

#bind new data into the set with princess anne/vb filtered out in relevant years 
newelectiondata <- bind_rows(elizabethcity, elizabethcityout)

```

#### Repeat process for Washabaugh and Jackson County, now just Jackson, SD. {style="color:teal;"}

```{r}
#repeat process for washabaugh-> jackson county SD
washabaugh <- newelectiondata %>% filter(GEO_ID %in% c("0500000US46131", "0500000US46071") & election_year %in% c(1924, 1928, 1932, 1936, 1940, 1944, 1948, 1952, 1956, 1960, 1964, 1968, 1972, 1976)) %>%
  mutate(GEO_ID= ifelse(county_name=="WASHABAUGH", "0500000US46071", GEO_ID), county_name= ifelse(county_name=="WASHABAUGH", "JACKSON", county_name)) %>%
  group_by(county_name, election_year, GEO_ID) %>% 
  summarize(across(c(democratic_raw_votes, republican_raw_votes, pres_raw_county_vote_totals_two_party), sum, na.rm=TRUE), .groups= 'drop')

#filter out in relevant years 
washabaughfilterout <- newelectiondata %>%
  filter(!(GEO_ID %in% c("0500000US46131", "0500000US46071") & election_year %in% c(1924, 1928, 1932, 1936, 1940, 1944, 1948, 1952, 1956, 1960, 1964, 1968, 1972, 1976)))

#bind new data into the set with Washabaugh/Jackson filtered out in relevant years 
newelectiondata <- bind_rows(washabaugh, washabaughfilterout)
```

#### Repeat process for Princess Anne and Virginia Beach, now just Virginia Beach, VA. {style="color:teal;"}

```{r}
#princess anne, va-> virginia beach

#change princess anne to virginia beach in years outside overlapping years 
namechange<- newelectiondata %>% filter(GEO_ID== "0500000US51151" & election_year %in% c(1916,1920,1924,1928,1932,1936,1940,1944,1948)) %>% mutate(GEO_ID= ifelse(county_name=="PRINCESS ANNE", "0500000US51810", GEO_ID), county_name= ifelse(county_name=="PRINCESS ANNE", "VIRGINIA BEACH", county_name)) 

#take overlapping years that virginia beach and princess anne existed
princessanne <- newelectiondata %>% filter(GEO_ID %in% c("0500000US51151", "0500000US51810") & election_year %in% c(1952, 1956, 1960)) %>%
  mutate(GEO_ID= ifelse(county_name=="PRINCESS ANNE", "0500000US51810", GEO_ID), county_name= ifelse(county_name=="PRINCESS ANNE", "VIRGINIA BEACH", county_name)) %>%
  group_by(county_name, election_year, GEO_ID) %>% 
  summarize(across(c(democratic_raw_votes, republican_raw_votes, pres_raw_county_vote_totals_two_party), sum, na.rm=TRUE), .groups= 'drop')

#filter out in relevant years 
princessanneout <- newelectiondata %>%
  filter(!(GEO_ID %in% c("0500000US51151", "0500000US51810") & election_year %in% c(1952, 1956, 1960)))

#bind new data into the set with princess anne/vb filtered out in relevant years 
newelectiondata <- bind_rows(princessanne, princessanneout, namechange)


```

#### Recheck again with changes that every election GEO_ID is in map set and vice versa. {style="color:teal;"}

```{r}
#| results: hide
#recheck
(newelectiondata$GEO_ID %>% unique)[!(newelectiondata$GEO_ID %>% unique) %in% (us_data2$GEO_ID %>% unique)]


#seems to double include princess anne, unrecognized GEOID. Drop this from our data


newelectiondata<- newelectiondata %>% filter(!(GEO_ID %in% c("0500000US51151", "0500000USNA")))

#recheck again- good now. 
(newelectiondata$GEO_ID %>% unique)[!(newelectiondata$GEO_ID %>% unique) %in% (us_data2$GEO_ID %>% unique)]

(us_data2$GEO_ID %>% unique)[!(us_data2$GEO_ID %>% unique) %in% (newelectiondata$GEO_ID %>% unique)]
```

#### Now that everything matches, let's make new variables in our election data set representing the percentage of Democrat and Republican vote, the dominant party (party who won each county), and by how much they won (the percent over 50%). {style="color:teal;"}

```{r}
#make new variables to give republican and democrat % of vote, as well as statements to take the republican/democrat vote if over 50% 
newelectiondata<- newelectiondata %>% 
  mutate(republican_raw_votes_2perc= republican_raw_votes/pres_raw_county_vote_totals_two_party) %>% mutate(democratic_raw_votes_2perc= 1-republican_raw_votes_2perc) %>% mutate(republican_excess=ifelse(republican_raw_votes_2perc>0.5, republican_raw_votes_2perc-0.5, 0), democrat_excess= ifelse(democratic_raw_votes_2perc>0.5, democratic_raw_votes_2perc-0.5, 0), dominantparty= ifelse(republican_raw_votes_2perc > democratic_raw_votes_2perc, "Republican", "Democrat"))




```

#### Now we can finally merge our map and election data on GEO_ID. {style="color:teal;"}

```{r}
#now lets finally merge the data 

mergedelectiondata<- full_join(newelectiondata, us_data2, by="GEO_ID")
```

##### Let's also create color scales for percentage of Republican and Democrat vote: {style="color:cornflowerblue;"}

```{r}
mapcolors <- mergedelectiondata %>% mutate(republicancolor = case_when(republican_excess>0.4 & republican_excess<=0.5 ~ "#a70000", republican_excess > 0.3 & republican_excess <= 0.4 ~ "#ff0000", republican_excess > 0.2 & republican_excess <= 0.3 ~ "#ff5252", republican_excess > 0.1 & republican_excess <= 0.2 ~ 	"#ff7b7b", republican_excess >= 0 & republican_excess<=0.1 ~ "#ffbaba", TRUE ~ NA_character_)) %>% mutate(democratcolor= case_when(democrat_excess>0.4 & democrat_excess<=0.5 ~ "#0200b9",democrat_excess > 0.3 & democrat_excess<= 0.4 ~ "#1000ff",democrat_excess > 0.2 & democrat_excess <= 0.3 ~"#0066cc" , democrat_excess > 0.1 & democrat_excess <= 0.2 ~"#3a80ec", democrat_excess >= 0 & democrat_excess<= 0.1 ~ "#cce0ff", TRUE ~ NA_character_)) 
```

#### Let's also rename us_data2 state data to actual state names for use later: {style="color:teal;"}

```{r}
us_data2<- us_data2 %>% mutate(STATE= case_when(
  STATE=="01"~"AL", 
  STATE=="04"~"AZ", 
  STATE=="05"~"AR", 
  STATE=="06"~"CA", 
  STATE=="08"~"CO", 
  STATE=="09"~"CT",
  STATE=="10"~"DE",
  STATE=="11"~"DC",
  STATE=="12"~"FL",
  STATE=="13"~"GA",
  STATE=="16"~"ID",
  STATE=="17"~"IL",
  STATE=="18"~"IN",       
  STATE=="19"~"IA",      
  STATE=="20"~"KS",
  STATE=="21"~"KY",
  STATE=="22"~"LA",
  STATE=="23"~"ME",      
  STATE=="24"~"MD",
  STATE=="25"~"MA",
  STATE=="26"~"MI",
  STATE=="27"~"MN",
  STATE=="28"~"MS",
  STATE=="29"~"MO",     
  STATE=="30"~"MT",
  STATE=="31"~"NE",       
  STATE=="32"~"NV",   
  STATE=="33"~"NH",       
  STATE=="34"~"NJ",       
  STATE=="35"~"NM",        
  STATE=="36"~"NY",      
  STATE=="37"~"NC",       
  STATE=="38"~"ND",        
  STATE=="39"~"OH",        
  STATE=="40"~"OK",       
  STATE=="41"~"OR",        
  STATE=="42"~"PA",       
  STATE=="44"~"RI",       
  STATE=="45"~"SC",      
  STATE=="46"~"SD",      
  STATE=="47"~"TN",       
  STATE=="48"~"TX",       
  STATE=="49"~"UT",       
  STATE=="50"~"VT",       
  STATE=="51"~"VA",       
  STATE=="53"~"WA",        
  STATE=="54"~"WV",     
  STATE=="55"~"WI",       
  STATE=="56"~"WY"))
```

#### Let's filter our data to just 2016 & 1916 to try two different individual election plots {style="color:teal;"}

```{r}
#filter to 2016 
year2016<- mapcolors %>% 
  filter(election_year==2016)
```

```{r}
year1916<- mapcolors %>% filter(election_year==1916)
```

## 2016 election plot: first, simply who won each county.

```{r}
ggplot(year2016) + geom_sf(aes(geometry=geometry, fill=ifelse(republican_raw_votes_2perc>0.5, "Republican", "Democrat"))) + scale_fill_manual(values= c("Republican"= "red", "Democrat"="blue"), name= "Winning Party") + ggthemes::theme_map()
```

#### This is a bit misleading so let's plot by how much they won: {style="color:teal;"}

```{r}
#plot for year 2016 dominant party in each county and by how much 
ggplot() + geom_sf(data= year2016 %>% filter(dominantparty=="Republican"), aes(geometry=geometry, fill=republicancolor), color="gray20", linewidth=0.085) + scale_fill_identity(name = "Republican Vote %", labels = c("90%-100%", "80%-90%", "70%-80%", "60%-70%", "50%-60%"), guide="legend") + new_scale_fill() + geom_sf(data= year2016 %>% filter(dominantparty=="Democrat"), aes(geometry=geometry, fill=democratcolor), color="gray20", linewidth=0.085) + scale_fill_identity(name= "Democrat Vote %", labels=c("90%-100%", "80%-90%", "70%-80%", "60%-70%", "50%-60%"), guide="legend") + ggthemes::theme_map() + theme(legend.position = "right")

```

## Try a different year, 1916:

```{r}
#plot for year 1916 who won
ggplot(year1916) + geom_sf(aes(geometry=geometry, fill=ifelse(republican_raw_votes_2perc>0.5, "Republican", "Democrat"))) + scale_fill_manual(values= c("Republican"= "red", "Democrat"="blue"), name= "Winning Party") + ggthemes::theme_map()
```

#### And again by how much in 1916: {style="color:teal;"}

```{r}
ggplot() + geom_sf(data= year1916 %>% filter(dominantparty=="Republican"), aes(geometry=geometry, fill=republicancolor), color="gray20", linewidth=0.085) + scale_fill_identity(name = "Republican Vote %", labels = c("90%-100%", "80%-90%", "70%-80%", "60%-70%", "50%-60%", "NA"), guide="legend") + new_scale_fill() +  geom_sf(data= year1916 %>% filter(dominantparty=="Democrat"), aes(geometry=geometry, fill=democratcolor), color="gray20", linewidth=0.085) + scale_fill_identity(name = "Democrat Vote %", labels = c("90%-100%", "80%-90%", "70%-80%", "60%-70%", "50%-60%", "NA"), guide="legend") + ggthemes::theme_map() + theme(legend.position = "right")


```

#### Arrange our data by year so that it is in chronological order: {style="color:teal;"}

```{r}
#arrange so that ggplot title works correctly 
mapcolors<- mapcolors %>% arrange(election_year)
```

#### Make a map for every single year (more efficiently in a for loop) and save it!: {style="color:teal;"}

```{r}
for (i in 1:27) {
  

current_year<- unique(mapcolors$election_year)[i]


g<-ggplot() + geom_sf(data= mapcolors %>% filter(dominantparty=="Republican"), aes(geometry=geometry, fill=republicancolor), color="gray20", linewidth=0.085) + scale_fill_identity(name = "Republican Vote %", labels = c("90%-100%", "80%-90%", "70%-80%", "60%-70%", "50%-60%"), guide="legend") + new_scale_fill() +  geom_sf(data= mapcolors %>% filter(dominantparty=="Democrat"), aes(geometry=geometry, fill=democratcolor), color="gray20", linewidth=0.085) + scale_fill_identity(name = "Democrat Vote %", labels = c("90%-100%", "80%-90%", "70%-80%", "60%-70%", "50%-60%"), guide="legend") + ggforce::facet_wrap_paginate(~election_year, ncol=1, nrow=1, page=i) + coord_sf(xlim = c(-140, -60)) + labs(title=paste("Election Year:", current_year)) + theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"), legend.position = "right", panel.background = element_blank(), axis.text = element_blank(),axis.ticks = element_blank(), axis.title = element_blank(), panel.grid = element_blank())


#ggsave(paste0("images/electionmap", i, ".png"), plot=g, width=10, height=8)

}


```

### Let's put all the plots together:

```{r}

facetwrap<-ggplot() + geom_sf(data= mapcolors %>% filter(dominantparty=="Republican"), aes(geometry=geometry, fill=republicancolor), color="gray20", linewidth=0.009) + scale_fill_identity(name = "Republican Vote %", labels = c("90%-100%", "80%-90%", "70%-80%", "60%-70%", "50%-60%"), guide="legend") + new_scale_fill() +  geom_sf(data= mapcolors %>% filter(dominantparty=="Democrat"), aes(geometry=geometry, fill=democratcolor), color="gray20", linewidth=0.009) + scale_fill_identity(name = "Democrat Vote %", labels = c("90%-100%", "80%-90%", "70%-80%", "60%-70%", "50%-60%"), guide="legend") + facet_wrap(~election_year, nrow=7, ncol=4) + coord_sf(xlim = c(-140, -60), expand = FALSE, clip = "off") + theme(panel.spacing = unit(0.1, 'lines'), plot.title = element_text(hjust = 0.5, size = 8, face = "bold"), legend.position = "right", panel.background = element_blank(), axis.text = element_blank(),axis.ticks = element_blank(), axis.title = element_blank(), panel.grid = element_blank())

facetwrap

#ggsave("facetwrap_plot.png", plot = facetwrap, width = 16, height = 12, dpi = 300)

```

### Let's visualize only more "current years": Bush, Clinton, Obama, Trump & Biden presidencies.

```{r}
#visualizing current years 
currentyears<- mapcolors %>% filter(election_year>=1988)

facetwrap2<-ggplot() + geom_sf(data= currentyears %>% filter(dominantparty=="Republican"), aes(geometry=geometry, fill=republicancolor), color="gray20", linewidth=0.01) + scale_fill_identity(name = "Republican Vote %", labels = c("90%-100%", "80%-90%", "70%-80%", "60%-70%", "50%-60%"), guide="legend") + new_scale_fill() +  geom_sf(data= currentyears %>% filter(dominantparty=="Democrat"), aes(geometry=geometry, fill=democratcolor), color="gray20", linewidth=0.01) + scale_fill_identity(name = "Democrat Vote %", labels = c("90%-100%", "80%-90%", "70%-80%", "60%-70%", "50%-60%"), guide="legend") + facet_wrap(~election_year, nrow=3, ncol=3) + coord_sf(xlim = c(-140, -60), expand = FALSE, clip = "off") + theme(plot.title = element_text(hjust = 0.5, size = 8, face = "bold"), legend.position = "right", panel.background = element_blank(), axis.text = element_blank(),axis.ticks = element_blank(), axis.title = element_blank(), panel.grid = element_blank())

facetwrap2

#ggsave("currentyears_plot.png", plot = facetwrap2, width = 16, height = 12, dpi = 300)
```

### Let's look only at the Republican vote over time:

```{r}
facetwrap3<-ggplot() + geom_sf(data= mapcolors %>% filter(dominantparty=="Republican"), aes(geometry=geometry, fill=republicancolor), color="gray20", linewidth=0.009) + scale_fill_identity(name = "Republican Vote %", labels = c("90%-100%", "80%-90%", "70%-80%", "60%-70%", "50%-60%"), guide="legend") + facet_wrap(~election_year, nrow=7, ncol=4) + coord_sf(xlim = c(-140, -60), expand = FALSE, clip = "off") + theme(panel.spacing = unit(0.1, 'lines'), plot.title = element_text(hjust = 0.5, size = 8, face = "bold"), legend.position = "right", panel.background = element_blank(), axis.text = element_blank(),axis.ticks = element_blank(), axis.title = element_blank(), panel.grid = element_blank())

facetwrap3
```

### And again only for current years:

```{r}
facetwrap4<-ggplot() + geom_sf(data= currentyears %>% filter(dominantparty=="Republican"), aes(geometry=geometry, fill=republicancolor), color="gray20", linewidth=0.01) + scale_fill_identity(name = "Republican Vote %", labels = c("90%-100%", "80%-90%", "70%-80%", "60%-70%", "50%-60%"), guide="legend") + facet_wrap(~election_year, nrow=3, ncol=3) + coord_sf(xlim = c(-140, -60), expand = FALSE, clip = "off") + theme(plot.title = element_text(hjust = 0.5, size = 10, face = "bold.italic", family="AppleGothic"), plot.subtitle = element_text(hjust = 0.5, size = 8, face = "bold.italic", family="Helvetica"), legend.title = element_text(size = 10, face="italic", family="Helvetica"), legend.key.size = unit(0.3, "cm"), panel.background = element_blank(), axis.text = element_blank(),axis.ticks = element_blank(), axis.title = element_blank(), panel.grid = element_blank(), strip.text = element_text(size = 8, face="bold"), strip.background = element_blank(), aspect.ratio = 0.35, panel.spacing = unit(0.2, "lines")) + labs(title="Republican Vote Distribution Across U.S. Counties in Modern Years", subtitle="Increasing both in size and strength")+ ggthemes::theme_hc()

facetwrap4

```

### Let's now make it interactive, starting with one year.

```{r}
library(jsonlite)
library(geojsonio)

#create simpler categories
highchartest<-mapcolors %>% filter(election_year==2016)
highchartest<- highchartest %>% mutate(color= case_when(dominantparty=="Republican" ~ republicancolor, dominantparty=="Democrat" ~ democratcolor), winningpercentage=ifelse(dominantparty=="Republican", republican_raw_votes_2perc*100, democratic_raw_votes_2perc*100))

#Transform shp to geojson 
mapjson<- geojson_json(us_data2)

#read geojson in 
Myhcmap <- jsonlite::fromJSON(mapjson, simplifyVector = FALSE)


#create tooltip table to incorporate all elements we want. 
x<-c("County Name:", "State:", "Winning Party:", "Winning Party's % of overall vote: ")
y<- c("{point.NAME}", "{point.properties.STATE}", "{point.dominantparty}", "{point.value:,.2f}%") #round to 2decimal points
tt<- tooltip_table(x,y)


#create html manually of labels for interactive map 
svg_content<- 
  "<svg>
    <rect x='10' y='40' width='35' height='20' stroke='black' fill='#cce0ff' stroke-width='3'/>
    <rect x='45' y='40' width='35' height='20' stroke='black' fill='#3a80ec' stroke-width='3'/>
    <rect x='80' y='40' width='35' height='20' stroke='black' fill='#0066cc' stroke-width='3'/>
    <rect x='115' y='40' width='35' height='20' stroke='black'fill='#1000ff' stroke-width='3'/>
    <rect x='150' y='40' width='35' height='20' stroke='black'fill='#0200b9' stroke-width='3'/>
    <text x='10' y='72' font-size='4' text-anchor='middle' fill='black' font-weight='bold'>50%</text>
    <text x='45' y='72' font-size='4' text-anchor='middle' fill='black' font-weight='bold' >60%</text>
    <text x='80' y='72' font-size='4' text-anchor='middle' fill='black' font-weight='bold'>70%</text>
    <text x='115' y='72' font-size='4' text-anchor='middle' fill='black' font-weight='bold'>80%</text>
    <text x='150' y='72' font-size='4' text-anchor='middle' fill='black' font-weight='bold'>90-100%</text>
    
    <text x='10' y='34' font-size='13' text-anchor='middle' fill='black' font-weight='bold' font-style='italic' font-family='Georgia, serif'>Democrat Vote %:</text>
    <rect x='10' y='100' width='35' height='20' stroke='black'fill='#ffbaba' stroke-width='3'/>
    <rect x='45' y='100' width='35' height='20' stroke='black'fill='#ff7b7b' stroke-width='3'/>
    <rect x='80' y='100' width='35' height='20' stroke='black'fill='#ff5252' stroke-width='3'/>
    <rect x='115' y='100'width='35' height='20' stroke='black'fill='#ff0000' stroke-width='3'/>
    <rect x='150' y='100'width='35' height='20' stroke='black'fill='#a70000' stroke-width='3'/>
    <text x='10' y='132' font-size='4' text-anchor='middle' fill='black' font-weight='bold'>50%</text>
    <text x='45' y='132' font-size='4' text-anchor='middle' fill='black' font-weight='bold' >60%</text>
    <text x='80' y='132' font-size='4' text-anchor='middle' fill='black' font-weight='bold'>70%</text>
    <text x='115' y='132' font-size='4' text-anchor='middle' fill='black' font-weight='bold'>80%</text>
    <text x='150' y='132' font-size='4' text-anchor='middle' fill='black' font-weight='bold'>90-100%</text>
    <text x='10' y='94' font-size='13' text-anchor='middle' fill='black' font-weight='bold' font-style='italic' font-family='Georgia, serif' >Republican Vote %:</text>
</svg>"


#make map
highchart(type = "map") %>%
   hc_add_series_map(df= highchartest, map= Myhcmap, name= "% of vote", value="winningpercentage", joinBy= "GEO_ID", showInLegend = T, borderColor="black", borderWidth=0.5) %>% hc_tooltip(pointFormat=tt, useHTML=TRUE) %>% hc_colorAxis(min=50,max=100) %>% hc_mapNavigation(enabled = TRUE) %>% hc_title(text= "2016 Election Results", style = list(color = "black", fontWeight = "bold")) %>% hc_subtitle(text="Winning Party's Share of Vote by County") %>% hc_caption(text=svg_content, align= "left", useHTML=TRUE) %>% hc_legend(enabled=FALSE) 

                                                                                                
```
